In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-20 paper/supplementary_main.tex

@@

\subsubsection{Steady flow in perivascular networks induced by pressure differences}

@@

\newpage
\section{Notes}

Various notes associated with \cite{eide2024functional}:
\begin{itemize}
  \item
    Can observations of tracer indicate compartmentalization without
    the existence of physical barriers but in the presence of
    e.g.~substantial flow? We made observations in this direction in
    \cite{vinje2021brain}, relating to the then-active
    shape-of-perivascular-spaces discussion. We can address this by
    trying to quantify and evaluate the permeability of the PVS-SAS
    interface - comparing model predictions with the timings and
    patterns presented by \cite{eide2024functional}.
  \item
    Do the tracer primarily move from the PVS into the SAS or along
    (more slowly) along the SAS? This is left as an open question by
    \cite{eide2024functional} (page 3)
  \item
    How do altered ICP affect transport? Note that PVSAS transport is
    impaired (delayed) with increasing mean wave amplitude in the ICP
    signal and for iNPH patients (with enlarged PVSAS spaces).
  \item
    Very useful reference for PVS sizes (10--20 mm$^2$ in reference
    patients, 15-70 mm$^2$ in iNPH), and average propagation speeds.
\end{itemize}

>>>>

\subsection*{Overview of in-silico scenarios and experiments considered}

See~\Cref{tab:scenarios} for an overview of the different in-silico
scenarios and computational models considered.
\begin{table}
  \begin{center}
  \begin{tabular}{ll|llllll|ll}
    \toprule
    & Description & $R_1, R_2$ & $\xi_{\rm PVS-SAS}$ &  $\beta_{\rm efflux}$ & $D$ & $\hat{u}$ & $\mathbf{u}_{SAS}$ & $c_0$ \\
    \midrule
    A & Baseline with sheaths & $R_2 = 2 R_1$ & $\xi$ & $10^{-4}$ mm$^2$/s\cite{hornkjol2022csf} & $\phi R D^{\rm gad}_{\rm eff}$\cite{sykova2008diffusion, valnes2020apparent} & $\hat{u}_{\rm prod} + \hat{u}_{\rm cardiac}$ & $\mathbf{u}_{\rm prod}$ & 0 \\ 
    B1 & Sheath permeability & $R_2 = 2 R_1$ & $0, \xi, 10 \xi, 100 \xi, 1000 \xi$ & $10^{-4}$ mm$^2$/s\cite{hornkjol2022csf} & $\phi R D^{\rm gad}_{\rm eff}$\cite{sykova2008diffusion, valnes2020apparent} & $\hat{u}_{\rm prod} + \hat{u}_{\rm cardiac}$ & $\mathbf{u}_{\rm prod}$ & 0 \\ 
    B2 & Health permeability & $R_2 = 2 R_1$ & $0, \xi, 10 \xi, 100 \xi, 1000 \xi$ & $10^{-4}$ mm$^2$/s\cite{hornkjol2022csf} & $\phi R D^{\rm gad}_{\rm eff}$\cite{sykova2008diffusion, valnes2020apparent} & $\hat{u}_{\rm prod} +  \hat{u}_{\rm cardiac} + \hat{u}_{\rm vaso}$ & $\mathbf{u}_{\rm prod}$ & 0 \\ 
    %C & Larger PVS & $R_2 = N R_1$ & -- & -- & --  &  --  & -- &  -- \\
    C & Higher PVS flow & -- & $\xi$ & -- & --  & $\hat{u}_{\rm prod} +  \hat{u}_{\rm cardiac} + \hat{u}_{\rm vaso}$  & -- & --  \\
    AA & PVS size & $R_2 = x \beta R_1$ & $\xi$ & -- & -- & -- &  -- & 0 \\ 
    %C & Larger PVS & $R_2 = N R_1$ & -- & -- & --  &  --  & -- &  -- \\
    B & PVS flow & -- & $\xi$ & -- & --  & $\hat{u}_{\rm prod} + \hat{u}_{\rm vaso}$  & -- & --  \\
    C & Clearance & -- & -- & -- & $\alpha D^{\rm gad, \tau, A\beta}$  & $\hat{u}_{\rm prod}$ & -- & 1 \\
    %E & PVS II & $R_2 = 2 R_1$ & $\xi$, -- & -- & --  &  $\hat{u} = \hat{u}_{\rm prod} + \uparrow\uparrow$  & -- & --  \\
    %F & PVS III & $R_2 = 2 R_1$ & $\xi$, -- & -- & $10 D_{\rm PVS}$ &  $\hat{u} = \hat{u}_{\rm prod}$  & -- & --  \\
    %G & Glymphatics & $R_2 = 2 R_1$ & $\xi$, -- & -- & \cite{sykova2008diffusion, valnes2020apparent} &  $\hat{u} = \hat{u}_{\rm prod} + \uparrow$  & $\mathbf{u}_{\rm PAR}$ > 0 & --  \\
    \bottomrule
    \end{tabular}
    \end{center}
  \caption{Overview of models (-- denotes as immediately above). Each
    PVS cross-section is modelled as a concentric annulus with inner
    radius $R_1$ and outer radius $R_2$ and such that $R_2 = \beta
    R_1$ with $\beta = 2.0$ based on~\cite{eide2024functional}. Let
    $\xi = \dots$ denote the permeability estimated by Koch et
    al.~\cite{koch2023estimates}. Model A and its variations aims to
    quantify the effect of PVS sheaths by varying the permeability
    $\xi$ between the PVS and its surroundings. We consider a fixed
    permeability $\xi_{\rm PVS-PAR}$ based
    on~\cite{koch2023estimates}. \mer{The permeability between the PVS
      and the SAS $\xi_{\rm PVS-SAS}$ must likely be estimated by a
      bit of trial-error from the observations of timings of PVS-SAS
      and SAS from~\cite{eide2024functional}.} \discuss{The
      extracranial solute efflux permeability $\beta_{\rm efflux}$ is
      uniformly distributed over the outer SAS boundary with a
      reasonable rate as baseline\cite{hornkjol2022csf,
        eide2021clinical, ringstad2024glymphatic}.} At baseline, the
    effective diffusion coefficients $D^{\rm x}_{\rm SAS} = D^{\rm
      x}_{\rm PVS}$ of the SAS and PVS represent the free diffusion
    coefficient of Gadubutrol (${\rm x} = {\rm gad}$) in CSF (water),
    and $D^{\rm x}_{\rm PAR}$ the effective diffusion coefficient of
    human cortical tissue, all at body temperature; weighted by a
    dispersion factor $\alpha$ (see Methods). At baseline, we consider
    net flow due to CSF-production in the SAS and PVS:
    $\mathbf{u}_{\rm SAS} > 0$ and $\hat{u} > 0$ (see Methods) but no
    bulk flow in the parenchyma $\mathbf{u}_{\rm PAR} = 0$. Due to the
    timescale for human intracranial tracer transport (minutes to
    hours) compared to typical (cardiac, respiratory, slow vasomotion)
    human intracranial pulsatility (0.1 -- 1 Hz), we consider net
    (constant-in-time) flow fields only in subsequent model variations
    (but see dispersion factor above).  Model B represents
    perivascular pathway scenarios with more rapid flow in the PVS
    induced by vasomotion (PVS I)~\cite{gjerde2023directional}.
    Models C represents a version of the aforementioned scenarios but
    for studying metabolite clearance (instead of solute influx).}
\label{tab:scenarios}
\end{table}

@@

\subsubsection{Steady flow in perivascular networks induced by pressure differences}
\label{sec:app:hydraulic_network}

@@
